---
title: "Code"
author: "Mahima Das"
format: html
embed-resources: true
editor: visual
---

```{r}
#load libraries

#| label: setup
#| message: false
library(easystats)
library(tidyverse)
library(brms)

baseline_year   <- 2002   
breakpoint_year <- 2014   
window_years    <- c(2002, 2026)

#'other (please specify)' options are scored and counted in the denominators and set to FALSE to score only the named response options
include_other_options <- TRUE

max_overrides <- list(
  construction   = NA,   
  participants   = NA,
  validity       = NA,
  reliability    = NA,
  efa            = NA,
  pca            = NA,
  cfa            = NA
)


#data is loaded here
codeframe_tib <- readr::read_csv(here::here("data/data.csv"), name_repair = "minimal", col_types = readr::cols(.default = readr::col_character())) |> 
  dplyr::slice(-1, -2)

#article metadata
meta_tib <- readr::read_csv(here::here("data/meta_data_final_dataset.csv")) |> 
  dplyr::mutate(doi = str_to_lower(str_trim(doi))) |> 
  dplyr::select(doi, publication_year)

```

## Screening the coded responses

```{r}
#| label: screening

#tick-box answers come out as text like "1,3" if more than one box was ticked,so check if a code is in the cell rather than testing it equals the code

selected <- function(x, codes) {
  str_split(as.character(x), ",\\s*") |>
    map_lgl(\(cd) any(cd %in% as.character(codes)))
}

#keep only finished responses for papers that could be open with nothing to exclude
screened_tib <- codeframe_tib |>
  dplyr::filter(
    selected(Finished, 1),
    selected(PaperAccess, 1),
    str_trim(as.character(PaperType)) == "1"   #no exclusion criteria apply
  ) |>
  dplyr::mutate(doi = str_to_lower(str_trim(SelectedItem)))

#for when papers are coded more than 1
duplicate_dois <- screened_tib |>
  count(doi) |>
  dplyr::filter(n > 1) |>
  pull(doi)

screened_tib <- screened_tib |>
  arrange(doi, StartDate) |>
  distinct(doi, .keep_all = TRUE)

#sanitycheck
tibble(
  responses_exported = nrow(codeframe_tib),
  double_coded       = length(duplicate_dois),
  papers_analysed    = nrow(screened_tib)
) |>
  display()
```

## Reshaping to one row per measure

```{r}
#| label: reshape-measures


#the export has a block of columns per measure, prefixed 1_ to 10_ thus pulling the numbers out rather than typing 1:10 in case an export has more
measure_blocks <- names(screened_tib) |>
  str_extract("^\\d+(?=_)") |>
  discard(is.na) |>
  unique() |>
  as.integer() |>
  sort()

#stacking the blocks so each measure gets its own row, prefix stripped off the column names so all blocks end up with matching names
measure_tib <- measure_blocks |>
  map(\(i) {
    pfx  <- paste0(i, "_")
    cols <- str_subset(names(screened_tib), paste0("^", pfx))
screened_tib |>
      dplyr::select(ResponseId, CoderID_1, doi, NumOfMeasures, all_of(cols)) |>
      rename_with(\(x) str_remove(x, fixed(pfx)) |> str_replace_all("[^A-Za-z0-9_]", "_"),
all_of(cols)) |>
      dplyr::filter(!is.na(MeasureName), str_trim(MeasureName) != "") |>
      dplyr::mutate(measure_num = i, .after = doi)
  }) |>
  list_rbind()

#sanitycheck
tibble(
  measure_blocks_found = length(measure_blocks),
  measures             = nrow(measure_tib),
  articles             = n_distinct(measure_tib$doi)
) |>
  display()
```

## Scoring the coding frame

```{r}
#| label: scoring-functions

#to count an item as "reported" if coded 1 or 2 (main text or supplementary)
reported     <- function(x) as.integer(selected(x, c(1, 2)))

#same idea but also allows code 3 as "reported" - used for the NumFactors/RotationMethod items where 4 means not reported
reported_123 <- function(x) as.integer(selected(x, c(1, 2, 3)))

#turns the item depiction question into 0/1/2 instead of raw qualtrics codes 2 = all items shown, 1 = just example items, 0 = nothing shown
item_depict  <- function(x) dplyr::case_when(
  selected(x, c(1, 2)) ~ 2L, #all items shown
  selected(x, c(3, 4)) ~ 1L,   #example items shown
  .default = 0L
)
is_other_option <- function(tib, cols) paste0(cols, "_TEXT") %in% names(tib) # to check whether a column has a matching "_TEXT" free-text column next to it

matching_cols <- function(tib, patterns) {
  cols <- names(tib) |>
    keep(\(nm) any(str_detect(nm, patterns))) |>
    discard(\(nm) str_ends(nm, "_TEXT"))
  if (!include_other_options) cols <- cols[!is_other_option(tib, cols)]
  cols
}

#scoring a set of columns and report the maximum those columns could reach
score_items <- function(tib, patterns, scorer = reported, max_per_item = 1) {
  cols <- matching_cols(tib, patterns)
  if (length(cols) == 0) {
    return(list(score = rep(0L, nrow(tib)), max = 0))
  }
  scores <- tib |>
    dplyr::select(all_of(cols)) |>
    dplyr::mutate(across(everything(), scorer))
  list(score = rowSums(scores), max = length(cols) * max_per_item)
}

#combine several scored blocks into one component
#e.g. combine_items(construction_part1, construction_part2) adds their scores together and adds their maxes together too
combine_items <- function(...) {
  parts <- list(...)
  list(
    score = reduce(map(parts, "score"), `+`),
    max   = sum(map_dbl(parts, "max"))
  )
}

#apply an override from the settings chunk if one was given- to manually fix a component's max (e.g. construction = 14 not 17)
apply_override <- function(part, name) {
  override <- max_overrides[[name]]
  if (!is.null(override) && !is.na(override)) part$max <- override
  part
}
```

```{r}
#| label: score-components

#descriptive information about scale construction (Q7-Q11)
construction <- combine_items(
  score_items(measure_tib, c("^ConstructDesc$", "^TargetPopDesc$", "^MeasureDevDesc$", "^MeasureDevDescWhat_\\d+$", "^ItemTotal$",           "^ResponseInstructions$", "^ScaleResponseFormat$",                           "^RatingScale$", "^ScoringProcedures$", "^ScoreInterpret$")),
  score_items(measure_tib, "^ItemDepict$", scorer = item_depict, max_per_item = 2)
) |>
  apply_override("construction")

#participant descriptions and missing data (Q12-Q14)
participants <- score_items(
  measure_tib,
  c("^NumParticipants$", "^NumGender$", "^NumAge$", "^ParticipantRecruit$",
    "^DataCollect_\\d+$", "^MissingDat$")
) |>
  apply_override("participants")

#validity (Q15.4 and Q15.5): one point each for reporting anything
convergent   <- score_items(measure_tib, "^Convergent_validity_\\d+$")
discriminant <- score_items(measure_tib, "^Discriminant_validit_\\d+$")
validity <- list(
  score = as.integer(convergent$score > 0) + as.integer(discriminant$score > 0),
  max   = 2
) |>
  apply_override("validity")

#reliability (Q15.3): internal consistency types
reliability <- score_items(measure_tib, "^InternalConsistency_\\d+$") |>
  apply_override("reliability")

#dimensionality: separate scores for EFA, PCA and CFA
efa <- combine_items(
  score_items(measure_tib, c("^NumFactorsEFA$", "^RotationMethodEFA$"), scorer = reported_123),
  score_items(measure_tib, "^EFADetail_\\d+$")
) |>
  apply_override("efa")

pca <- combine_items(
  score_items(measure_tib, c("^NumFactorsPCA$", "^RotationMethodPCA$"), scorer = reported_123),
  score_items(measure_tib, "^PCADetail_\\d+$")
) |>
  apply_override("pca")

cfa <- combine_items(
  score_items(measure_tib, "^CFADetail_\\d+$"),
  score_items(measure_tib, "^CFAMods$")
) |>
  apply_override("cfa")

#sanitycheck
tibble(
  component = c("construction", "participants", "validity", "reliability",
                "efa", "pca", "cfa"),
  maximum   = c(construction$max, participants$max, validity$max, reliability$max,
                efa$max, pca$max, cfa$max)
) |>
  display()
```

```{r}
#| label: build-scores

scored_tib <- measure_tib |>
  dplyr::mutate(
    construction      = construction$score,
    participants_raw  = participants$score,
    validity          = validity$score,
    reliability       = reliability$score,
    
#which dimensionality analyses were reported for this measure
    has_efa = reported(Dimensonality_1),
    has_cfa = reported(Dimensonality_2),
    has_pca = reported(Dimensonality_3),
    
    #efa and pca scored, each is 0 out for anyone who didn't report that particular analysis
    efa_score = efa$score * has_efa,
    efa_denom = efa$max * has_efa,
    pca_score = pca$score * has_pca,
    pca_denom = pca$max * has_pca,
    cfa_score = cfa$score * has_cfa,
    cfa_denom = cfa$max * has_cfa
  )

#if SameSample == 1 the participant block was not shown again, so the previous measure's participant information applies to the current measure
scored_tib <- scored_tib |>
  arrange(ResponseId, measure_num) |>
  dplyr::mutate(
    participants = if_else(selected(SameSample, 1), NA_real_, participants_raw)
  ) |>
  group_by(ResponseId) |>
  fill(participants, .direction = "down") |>
  ungroup() |>
  dplyr::mutate(participants = replace_na(participants, 0))
```

```{r}
#| label: proportions

construction_max <- construction$max
participants_max <- participants$max
validity_max     <- validity$max
reliability_max  <- reliability$max


scored_tib<- scored_tib |>
  dplyr::mutate(
    #overall score (primary outcome)
    total_score = construction + participants + validity + reliability +
      efa_score + pca_score + cfa_score,
    total_max   = construction_max + participants_max + validity_max +
      reliability_max + efa_denom + pca_denom + cfa_denom,
    prop_score  = total_score / total_max,
    
    #sub-dimension proportions for the secondary analyses
    prop_construction = construction / construction_max,
    prop_participants = participants / participants_max,
    prop_valrel       = (validity + reliability) / (validity_max + reliability_max), prop_dimensionality = if_else(
      efa_denom + pca_denom + cfa_denom > 0,
      (efa_score + pca_score + cfa_score) / (efa_denom + pca_denom + cfa_denom),
      NA_real_
    )
  )
```

```{r}
#| label: analysis-data


analysis_tib <- scored_tib |>
  left_join(meta_tib, by = "doi") |>
  dplyr::mutate(time = publication_year - baseline_year,
                era = as.integer(publication_year >= breakpoint_year),
                time_post = (time - (breakpoint_year - baseline_year)) * era
  )
#sanity check
tibble(
  measures      = nrow(analysis_tib),
  articles      = n_distinct(analysis_tib$doi),
  unmatched_doi = sum(is.na(analysis_tib$publication_year)),
  out_of_window = sum(!between(analysis_tib$publication_year,
                               window_years[1], window_years[2]), na.rm = TRUE)
) |>
  display()
analysis_tib <- analysis_tib |>
  dplyr::filter(!is.na(publication_year),
                between(publication_year, window_years[1], window_years[2]))
```

## Summary statistics

```{r}
#| label: descriptives

#reset random number generator for reproducibility
set.seed(1234)

#the outcomes carried through to the models
outcomes <- c(
  overall        = "prop_score",
  construction   = "prop_construction",
  participants   = "prop_participants",
  validity_rel   = "prop_valrel",
  dimensionality = "prop_dimensionality"
)
#summary stats (mean, sd etc) for every outcome at once
analysis_tib |>
  dplyr::select(all_of(unname(outcomes))) |>
  describe_distribution() |>
  data_remove(c(Kurtosis, Skewness)) |>
  display(digits = 3)

#how many measures sit exactly on 0 or 1, which decides whether each outcome is modelled with a beta or a zero-one-inflated beta likelihood
analysis_tib |>
  dplyr::select(all_of(unname(outcomes))) |>
  pivot_longer(everything(), names_to = "outcome", values_to = "p") |>
  dplyr::filter(!is.na(p)) |>
  group_by(outcome) |>
  summarise(n = n(), zeros = sum(p == 0), ones = sum(p == 1), .groups = "drop") |>
  display()
```

```{r}
#| label: fig-measures-per-article

#the clustering the random intercept is there to handle
measures_plot <- analysis_tib |>
  count(doi, name = "measures") |>
  ggplot(aes(x = measures)) +
  geom_bar(fill = "#999933", colour = "black", alpha = 0.7) +
  scale_x_continuous(breaks = scales::breaks_width(1)) +
  labs(x = "Measures coded within an article", y = "Number of articles") +
  theme_minimal()

measures_plot
```

## Primary analysis

The outcome is the proportion of coding scheme items reported for each measure. Measures are nested within articles, so the model carries a random intercept for article:

$$\begin{aligned}
g\left(\mu_{ij}\right) &= \ln\left(\frac{\mu_{ij}}{1-\mu_{ij}}\right) = \eta_{ij}\\
\eta_{ij} &= \left[\beta_{0} + \beta_1\text{time}_{i} + \beta_2\left(\text{time}_{i}-12\right)\times\text{era}_{i}\right] + \left[\mu_{0j} + \varepsilon_{ij}\right]
\end{aligned}$$

$\beta_1$ is the slope pre-2014 and $\beta_2$ is the *difference* in slope post-2014, so the post-2014 slope is $\beta_1 + \beta_2$. A beta likelihood cannot accommodate proportions of exactly 0 or 1, so a zero-one-inflated beta is fitted instead whenever an outcome touches those boundaries.

```{r}
#| label: model-function


fit_reporting_model <- function(data, outcome, slope_prior, prior_only = FALSE) {
  model_tib <- data |>
    dplyr::filter(!is.na(.data[[outcome]])) |>
    dplyr::rename(p = all_of(outcome))
  
  model_family <- if (any(model_tib$p %in% c(0, 1))) {
    brms::zero_one_inflated_beta()
  } else {
    brms::Beta(link = "logit")
  }
  
  f <- brms::brmsformula(p ~ 0 + Intercept + time + time_post + (1 | doi))
  
  model_priors <- c(
    slope_prior,
    brms::set_prior("normal(0, 3)", class = "b", coef = "Intercept")
  )
  
  brms::brm(
    formula      = f,
    data         = model_tib,
    family       = model_family,
    prior        = model_priors,
    sample_prior = if (prior_only) "only" else "no",
    chains       = 4,
    iter         = 8000,
    cores        = 4,
    seed         = 1234,
    refresh      = 100,
    backend= "cmdstanr",
    control      = list(adapt_delta = 0.95)
  )
}
```

```{r}
#| label: prior-check
#| message: false

set.seed(1234)
#dont run-IGNORE
overall_prior_brm <- fit_reporting_model(
  data        = analysis_tib,
  outcome     = "prop_score",
  slope_prior = brms::set_prior("normal(0.5, 0.25)", class = "b"),
  prior_only  = TRUE
)

prior_pred_tib <- estimate_relation(overall_prior_brm, keep_iterations = 100) |>
  reshape_iterations()

ggplot(prior_pred_tib, aes(x = time + baseline_year, y = iter_value, group = iter_group)) +
  geom_line(alpha = 0.15, colour = "#999933") +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "Publication year", y = "Prior-implied proportion") +
  theme_minimal()
```

```{r}
#| label: fit-primary
#| message: false

set.seed(1234)
#dont run-IGNORE

#reporting expected flat pre-2014, improving post-2014
overall_brm <- fit_reporting_model(
  data        = analysis_tib,
  outcome     = outcomes[["overall"]],
  slope_prior = brms::set_prior("normal(0.5, 0.25)", class = "b")
)
```

```{r}
#| label: primary-checks

#hairy caterpillars - one trace per chain, no pattern
plot(overall_brm)

#rhat should be < 1.01, ess > 1000
diagnostic_posterior(overall_brm) |> display(digits = 3)

#does the model reproduce the observed distribution?
check_predictions(overall_brm) |> plot()
```

```{r}
#| label: primary-parameters

#median point estimate, 95% hdi, pd (existence) and rope
model_parameters(
  overall_brm,
  centrality = "median",
  ci         = 0.95,
  ci_method  = "hdi",
  test       = c("pd", "rope"),
  effects    = "all"
) |>
  display(digits = 3)

#beta2 is the change in slope; the post-2014 slope is beta1 + beta2
hypothesis(overall_brm, "time_post > 0")
hypothesis(overall_brm, "time + time_post > 0")

#variance explained (interpret loosely - this is a glm, not linear)
performance::r2(overall_brm)
```

```{r}
#| label: fig-primary-trend

#predict the trend across the window, keeping iterations for the fan of lines
trend_wide <- tibble(time = seq(0, window_years[2] - baseline_year)) |>
  dplyr::mutate(
    era       = as.integer(time >= breakpoint_year - baseline_year),
    time_post = (time - (breakpoint_year - baseline_year)) * era,
    doi       = NA_character_               #population-level, ignore random effect
  ) |>
  estimate_relation(model = overall_brm, data = _, keep_iterations = 100) |>
  dplyr::mutate(year = time + baseline_year)

trend_long <- reshape_iterations(trend_wide)

ggplot(analysis_tib, aes(x = publication_year)) +
  geom_point(aes(y = prop_score), alpha = 0.3) +
  geom_line(data = trend_long, aes(x = year, y = iter_value, group = iter_group),
            alpha = 0.08, colour = "#999933") +
  geom_ribbon(data = trend_wide, aes(x = year, ymin = CI_low, ymax = CI_high),
              fill = "#999933", alpha = 0.25) +
  geom_line(data = trend_wide, aes(x = year, y = Predicted),
            colour = "#999933", linewidth = 0.9) +
  geom_vline(xintercept = breakpoint_year, linetype = "dashed") +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "Publication year", y = "Proportion of items reported") +
  theme_minimal()
```

```{r}
#| label: fit-secondary

#dont run-IGNORE
set.seed(1234)

secondary_brms <- outcomes[names(outcomes) != "overall"] |>
  map(\(outcome) {
    fit_reporting_model(
      data        = analysis_tib,
      outcome     = outcome,
      slope_prior = brms::set_prior("normal(0, 1)", class = "b")
    )
  })
```

```{r}
#| label: secondary-parameters

#posterior summaries for every sub-dimension in one table
secondary_brms |>
  imap(\(model, label) {
    model_parameters(model, centrality = "median", ci = 0.95,
                     ci_method = "hdi", test = c("pd", "rope")) |>
      as_tibble() |>
      dplyr::mutate(outcome = label, .before = 1)
  }) |>
  list_rbind() |>
  dplyr::filter(str_detect(Parameter, "Intercept|^time")) |>
  display(digits = 3)
```

```{r}
#| label: fig-secondary-trends

#observed sub-dimension proportions across time
analysis_tib |>
  dplyr::select(publication_year,
                all_of(unname(outcomes[names(outcomes) != "overall"]))) |>
  pivot_longer(-publication_year, names_to = "outcome", values_to = "p") |>
  dplyr::filter(!is.na(p)) |>
  group_by(publication_year, outcome) |>
  summarise(p = mean(p), n = n(), .groups = "drop") |>
  ggplot(aes(x = publication_year, y = p, colour = outcome)) +
  geom_line() +
  geom_point(aes(size = n), alpha = 0.6) +
  geom_vline(xintercept = breakpoint_year, linetype = "dashed") +
  coord_cartesian(ylim = c(0, 1)) +
  scale_colour_viridis_d(begin = 0.3, end = 0.85) +
  labs(x = "Publication year", y = "Mean proportion of items reported",
       colour = "Sub-dimension", size = "k measures") +
  theme_minimal()
```

```{r}
secondary_brms |>
  purrr::imap(\(m, nm) {
    parameters::model_parameters(m, centrality = "median", ci = 0.95,
                                 ci_method = "hdi", test = c("pd", "rope")) |>
      tibble::as_tibble() |>
      dplyr::mutate(outcome = nm, .before = 1)
  }) |>
  purrr::list_rbind() |>
  dplyr::filter(Parameter %in% c("b_time", "b_time_post"))
```

```{r}

#reading the already saved models
overall_brm    <- readRDS("D:/SUSSEX/mahima_das_2026_dissertation/models/overall_brm.rds")
secondary_brms <- readRDS("D:/SUSSEX/mahima_das_2026_dissertation/models/secondary_brms.rds")
```

```{r}
secondary_brms |>
  purrr::imap(\(m, nm) {
    parameters::model_parameters(m, centrality = "median", ci = 0.95,
                                 ci_method = "hdi", test = c("pd", "rope")) |>
      tibble::as_tibble() |>
      dplyr::mutate(outcome = nm, .before = 1)
  }) |>
  purrr::list_rbind() |>
  dplyr::filter(Parameter %in% c("b_time", "b_time_post"))
```

```{r}
#ignore-already saved
saveRDS(overall_brm, "D:/SUSSEX/mahima_das_2026_dissertation/models/overall_brm.rds")
saveRDS(secondary_brms, "D:/SUSSEX/mahima_das_2026_dissertation/models/secondary_brms.rds")
```

```{r}
#IGNORE
plot(parameters::model_parameters(overall_brm, effects = "fixed"))
```

